Left-linear rules in adhesive categories

### 2024-02-13 icalp/fusione.tex
typo
conclusioni

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagram mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.
%\todo{P: la sequenza di lavori su questo \`e ampia,
%vedere se c'\`e qualcosa d'altro}

@@

There is, however, a further problem.  For linear
rewriting systems, switch equivalence can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}.
%(roughly, an isomorphism between their colimits). 
This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

\todo{P: sopra era $X \in\Deltamin(\der{D})$, non definito nel main}

@@

The characterisation of switch equivalence based on consistent permutations is the key to
derive some simple but fundamental properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}

In particular, the two properties allows one to obtain a \emph{canonical form} for switching sequences: whenever two derivation sequences are switch equivalence via some permutation $\sigma$, then each one can be obtained from the other by applying switchings to pairs of rule applications which $\sigma$ reverses (the so-called inversions).\todo{P: un po' di questi discorsi sono copiati nell'intro, ridurre}

The failure of the characterisation of switch equivalence via
consistent permutations for left-linear rules leads to the
question asking whether the properties above hold.

In this section we show that actually much of the theory can be retained: confluence of switching still holds, while globality of independence holds in a weak form, conceptually linked to the fact that fusions introduce a form of disjunctive causality. Finally, a canonical form for switching sequences can also be recovered.

>>>>

So far, most of the theoretical results focussed on what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens already
in the context of string diagram mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

@@

% \begin{enumerate}
% \item \emph{confluence of switching}: when transforming derivation
%   sequences into equivalent ones by switching independent steps, the
%   result not depends on the order in which switchings are applied, but
%   only on the resulting permutation;

% \item \emph{globality of independence}: if two independent steps are
%   moved forward or backward due to the application of switchings to a
%   derivation, they remain independent.
% \end{enumerate}

There is, however, a further problem. In the literature on linear
rewriting systems, a central result is the a characterisation of
switch equivalence of derivations based on so-called processes or
consistent permutations~\cite{add citations}.\todo{Handboo, graph processes, tobias}
In turn this is the key
to derive some simple but fundamental properties for linear rewriting
systems, namely

\begin{enumerate}
\item \emph{consistency of switching}: when transforming
  sequences of rewriting steps into equivalent ones by switching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;
\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs 
between sequences of rewriting steps seen as concurrent computations.

@@

@@

As discussed in the introduction, a number of basic properties of
switch equivalence, i.e., the confluence of swicthing, globality of
independence and the existence of a canonical form for equivalence
proofs, can be derived, in the linear case, from a characterisation of
switch equivalence in terms of so-called consistent permutations or
processes.

When dealing with left-linear rules, such a characterisation fails. Leaving aside the formal details (the interested reader can consult Appendix~\ref{})\todo{P: possiamo mettere in appendice la def. consistent permutation? e forse anche l'implicazione che vale per i left linear?}, the intuition is quite simple. In the linear case one proves that two derivations is using the same rules in different orders are switch equivalent when the colimits of the two derivations, seen as diagrams, are isomorphic and the isomorphism properly commutes with the embeddings of the rules.

This result clearly fails when considering left-linear
derivations. This is readily seen by considering derivations $\der{D}$
in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
already observed that $\der{D} \not\equiv_a \der{D}'$\todo{P:
  osservato?}
% 
Still, it is easy to see that they are related by a consistent
permutation, which is the identity on productions. The colimit of
both derivations is a single node with a self loop, i.e., the
terminal object in $\cat{Graph}$, hence all commutation requirements
are trivially satisfied.

In this section we show, despite the failure of the s, actually much of the theory can
be retained: confluence of switching still holds, while globality of
independence holds in a weak form, conceptually linked to the fact
that fusions introduce a form of disjunctive causality. Finally, a
canonical form for switching sequences can also be recovered.
